Bumble bees and wildflowers along an elevational climate gradient exhibit niche

### 2020-07-03 analyses/preference.Rmd
fixed some things, broke some things

### Load network and survey data
```{r, echo=FALSE, include=FALSE}
### BB traits
bb_traits <- read_csv("../processed_data/bb_traits.csv") %>%
  dplyr::select(subgenus, bb.sp, pbl.w, pbl.w.class)

### FL Kugler classification
fl_traits <- read.csv("../processed_data/floral_k_type_nom_fix.csv") %>%
  dplyr::select(plant.sp, k.type) %>%
  mutate(k.type.s = str_extract(k.type, "[-+]?[0-9]*\\.?[0-9]+"),
         k.type.ss = str_extract(k.type, "[-+]?[0-9]*"))

### FL corolla depth (from Benadi)
fl_corolla_depth <- read.csv("../processed_data/flower_morphology_GitaBenadi.csv") %>%
  group_by(plant.sp = Species) %>%
  summarize(corolla.depth = mean(LN))

### FL higher  taxonomy
fl_tax <- read_csv("../processed_data/floral_tax.csv")

### Colony life cycle break points
breaks <- read_csv("./output/breaks.csv") %>%
  mutate(year = factor(year))

### BB-FL visitation data
net <- read_csv("../processed_data/network.csv") %>%
  mutate(year = factor(year)) %>% # convert year to factor
  full_join(bb_traits, by = c("bb.sp" = "bb.sp")) %>%
  left_join(site_data, by = c("site")) %>%
  left_join(gdd, by = c("site", "date", "year")) %>%
  left_join(breaks, by = c("site", "year")) %>%
  left_join(fl_traits) %>%
  left_join(fl_corolla_depth) %>%
  left_join(fl_tax, by = "plant.genus") %>%
  #left_join(temperature, by = c("site", "date")) %>%
  ungroup() %>%
  mutate(phase = case_when(
    gdd.cum < break1.gdd ~ "founding",
    gdd.cum >= break1.gdd & gdd.cum < break2.gdd ~ "buildup",
    gdd.cum > break2.gdd ~ "reproductive"
  )) %>%
  mutate(phase = factor(phase, levels = c("founding", "buildup", "reproductive"))) %>%
  mutate(caste = case_when(
    caste == "worker" ~ "worker",
    caste == "male" ~ "reproductive",
    caste == "youngqueen" ~ "reproductive",
    caste == "queen" ~ "foundress",
    caste == "oldqueen" ~ "foundress"
  )) %>%
  rename(yday = dayofyear) %>%
  arrange(site, date)

### Floral survey data
survey <- read_csv("../processed_data/floral_survey.csv") %>%
  mutate(year = factor(year)) %>% # convert year to factor
  semi_join(net, by = "plant.sp") %>% # Consider only plants visited (anytime, not specifically for a given site*date)
  left_join(gdd, by = c("site", "date", "year")) %>%
  left_join(breaks, by = c("site", "year")) %>%
  left_join(site_data, by = c("site")) %>%
  left_join(fl_traits) %>%
  left_join(fl_traits) %>%
  left_join(fl_tax) %>%
  #left_join(temperature, by = c("site", "date")) %>%
  ungroup() %>%
  mutate(phase = case_when(
    gdd.cum < break1.gdd ~ "founding",
    gdd.cum >= break1.gdd & gdd.cum < break2.gdd ~ "buildup",
    gdd.cum > break2.gdd ~ "reproductive"
  )) %>%
  mutate(phase = factor(phase, levels = c("founding", "buildup", "reproductive"))) %>%
  rename(yday = dayofyear) %>%
  arrange(site, date)
```

### FL and BB abundance / diversity
```{r echo=FALSE, include=FALSE}
### Per-transect BB abundance: species-level
bb_abund_sp <- net %>%
  group_by(site, date, yday, gdd.cum, year, elev.class2, 
           elev.mean, phase, break1.gdd, break2.gdd, bb.sp) %>%
  summarize(bb.abund = n()) %>%
  ungroup() %>%
  spread(bb.sp, bb.abund) %>%
  gather(bb.sp, bb.abund, -c(site, date, year, yday, gdd.cum, elev.class2, 
                             elev.mean, phase, break1.gdd, break2.gdd)) %>%
  replace_na(list(bb.abund = 0)) %>%
  mutate(bb.sp = factor(bb.sp))

### Per-transect BB abundance: species-level, by caste
bb_abund_sp_caste <- net %>%
  group_by(site, date, yday, gdd.cum, year, elev.class2, 
           elev.mean, phase, break1.gdd, break2.gdd, bb.sp, caste) %>%
  summarize(bb.abund = n()) %>%
  group_by(caste) %>%  # The next several lines of crazy nonsense are all about filling in zeros where a given caste*species was not observed.
  spread(bb.sp, bb.abund) %>%
  gather(bb.sp, bb.abund, -c(site, date, year, yday, gdd.cum, elev.class2, 
                             elev.mean, phase, break1.gdd, break2.gdd, caste)) %>%
  group_by(bb.sp) %>%
  spread(caste, bb.abund) %>%
  gather(caste, bb.abund, -c(site, date, year, yday, gdd.cum, elev.class2, 
                             elev.mean, phase, break1.gdd, break2.gdd, bb.sp)) %>%
  replace_na(list(bb.abund = 0)) %>%
  ungroup() %>%
  mutate(bb.sp = factor(bb.sp))

### Per-transect floral abund: by species
fl_abund_sp <- survey %>%
  group_by(site, date, year, plant.sp, yday) %>%
  summarize(fl.abund = sum(flower.cover)) %>%
  ungroup() %>%
  spread(plant.sp, fl.abund) %>%
  gather(plant.sp, fl.abund, -c(site, date, year, yday)) %>%
  replace_na(list(fl.abund = 0)) %>%
  mutate(plant.sp = factor(plant.sp)) %>%
  dplyr::select(site, year, yday, plant.sp, fl.abund) %>%  # this is an attempt to manage the memory crashes I'm getting with a later join step
  group_by(site, year, yday) %>%
  mutate(prop.fl.abund = fl.abund/sum(fl.abund))
```

### Get most important plants for each BB species
```{r}
# Per transect visitation rates of flower species per BB species
visit_rates <- net %>%
  group_by(bb.sp, plant.sp, year, site, yday) %>%
  summarize(visits = n()) %>%
  group_by(bb.sp) %>%
  spread(plant.sp, visits) %>%
  gather(plant.sp, visits, -c(bb.sp, year, site, yday)) %>%
  group_by(plant.sp) %>%
  spread(bb.sp, visits) %>%
  gather(bb.sp, visits, -c(plant.sp, year, site, yday)) %>%
  group_by(year, site, bb.sp, yday) %>%
  replace_na(list(visits = 0)) %>%
  group_by(site, year, yday, bb.sp) %>%
  mutate(prop.visits = visits/sum(visits)) %>% # This will create NAs wherever a given bb=fl combo was all zeros, so we need to set the NAs to 0 
  replace_na(list(prop.visits = 0)) %>%
  arrange(site, year, yday, bb.sp, -prop.visits) %>%
  ungroup() %>%
  dplyr::select(year, site, yday, bb.sp, plant.sp, visits, prop.visits) # this is an attempt to manage the memory crashes I'm getting with the subsequent join step

### Join FL abundance data to FL visitation data
fl_abund_x_visitation <- visit_rates %>%
  inner_join(fl_abund_sp) %>%
  filter(fl.abund > 0 & prop.fl.abund < 1) %>% # Can't calculate preference where the plant.sp is absent or where it is the only plant present
  inner_join(bb_abund_sp, by = c("site", "year", "yday", "bb.sp")) %>%
  filter(bb.abund > 0) # Can't calculate preference where the bb.sp is absent

fl_abund_x_visitation.wm <- fl_abund_x_visitation %>%
  group_by(year, bb.sp, plant.sp) %>%
  summarize(prop.fl.abund.m = mean(prop.fl.abund),
            prop.visits.wm = weighted.mean(prop.visits, bb.abund))
```

### Okay, here's what we're going to do. Preference has a strong bimodal pattern, with lots of plant species totally ignored while the ones that are foraged upon are visited in proportion to their abundance. We can't model that latter pattern well if we retain all the totally ignored plant species. So, let's model the visited plants separately from the unvisited ones. 
```{r}
### Call linear model on the subset of the floral species that were visited at least once
pref.lm <- lm(log(prop.visits.wm) ~ log(prop.fl.abund.m), # log on log regression is common for proportional data
              data = filter(fl_abund_x_visitation.wm, prop.visits.wm > 0))

@@

ggplot(filter(fl_abund_x_visitation.wm, prop.visits.wm > 0),
       aes(log(prop.fl.abund.m), log(prop.visits.wm))) +
  geom_point() +
  geom_smooth(method = "lm")

@@

Join residuals back into data and summarize by site-date-bb.sp (this deserves its own code chunk)
```{r}
#################################
### Build preference data set ###
#################################
# Start with the mean prop.fl.abund and weighted mean prop.visits
pref <- fl_abund_x_visitation.wm %>%  
  ungroup() %>%

@@

# Now we need to add back all the bb*fl*year combos that had 0 visits
# They we have NAs for preference and for floral abundance columns
  full_join(unique(dplyr::select(visit_rates, year, bb.sp, plant.sp))) %>%

@@

# Drop the columns used to calculate preference so that they
# don't confuse us when we add site-date level abundance data
  dplyr::select(year, bb.sp, plant.sp, preference)

@@

pref_wt_abund <- fl_abund_sp %>%
  full_join(pref, by = c("year", "plant.sp")) %>%  # This step introduces a small handful of NA because there were certain cases where a plant species was not observed at all in a given year. We will just omit those zeros from our data since they won't affect the analysis significantly
  mutate(fl.abund.weighted = fl.abund*preference) %>%
  group_by(site, year, yday, bb.sp) %>%
  summarize(fl.abund.weighted = sum(fl.abund.weighted, na.rm = TRUE)) %>%
  left_join(site_data, by = "site") %>%
  filter(!is.na(site)) %>%
  left_join(bb_abund_sp)
```

Preference-weighted floral abundance by year
```{r}
pref_sum <- pref_wt_abund %>%
  group_by(site, year, bb.sp) %>%
  summarize(fl.abund.weighted = mean(fl.abund.weighted))

ggplot(pref_sum, aes(year, fl.abund.weighted, color = bb.sp)) +
  geom_point()

ggplot(pref_sum, aes(year, log(fl.abund.weighted))) +
  geom_boxplot() +
  facet_wrap(~bb.sp, scales = "free")
```

GAM modeling of preference-weighted abundance per bb species
## Against any potential objections to my use of an added constant to avoid the log(0) problem, I would argue that when what is being measured is a resource, the presence of a resource in vanishingly small amounts is biologically equivalent to its absence. I suppose the strength of this argument hinges on how much we trust that our samples represent our sites. But if they don't, what are we even doing here? note that this argument does not hold when one is concerned with diversity and species distributions, since then a species rarity is a very different thing that its absence.
```{r}
### Set constants
k <- min(
  filter(pref_wt_abund, fl.abund.weighted > 0)$fl.abund.weighted
)

### Call models
weighted_fl_abundance.2010 <- bam(log(fl.abund.weighted + k) ~ te(yday, elev.mean, 
                                                  by = factor(bb.sp),
                                                  bs= c("tp", "tp"), 
                                                  k=c(10, 10), m=1),
                        data = filter(pref_wt_abund, year == 2010),
                        discrete = TRUE,
                        method = "fREML") %>% getViz()

@@

weighted_fl_abundance.2011 <- bam(log(fl.abund.weighted + k) ~ te(yday, elev.mean, 
                                                  by = factor(bb.sp),
                                                  bs= c("tp", "tp"), 
                                                  k=c(10, 10), m=1),
                        data = filter(pref_wt_abund, year == 2011),
                        discrete = TRUE,
                        method = "fREML") %>% getViz()

@@

weighted_fl_abundance.2012 <- bam(log(fl.abund.weighted + k) ~ te(yday, elev.mean, 
                                                  by = factor(bb.sp),
                                                  bs= c("tp", "tp"), 
                                                  k=c(10, 10), m=1),
                        data = filter(pref_wt_abund, year == 2012),
                        discrete = TRUE,
                        method = "fREML") %>% getViz()

>>>>

### Load network and survey data
```{r}
### BB-FL visitation data
net <- read_csv("../processed_data/network.csv") %>%
  dplyr::select(site, date, bb.sp, plant.sp)

### Floral survey data
survey <- read_csv("../processed_data/floral_survey.csv") %>%
  semi_join(net, by = "plant.sp") %>% # Consider only plants visited (anytime anywhere, not specifically for a given site*date)
  dplyr::select(site, date, plant.sp, flower.cover)

shared_transects <- inner_join(net, survey, by = c("site", "date")) %>%
  dplyr::select(site, date) %>%
  unique()
```

### FL and BB abundance / diversity
```{r echo=FALSE, include=FALSE}
### Per-transect BB abundance: species-level
bb_abund_sp <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  semi_join(shared_transects) %>% # I only want BB data with corresponding floral data
  group_by(site, date, bb.sp) %>% # group by bb.sp*transect (and other variables we want to retain)
  summarize(bb.abund = n()) %>% # get bb.sp*transect abundance
  ungroup %>%
  complete(site, date, bb.sp, fill = list(bb.abund = 0)) # complete implicit absences

### Per-transect floral abund: by species
fl_abund_sp <- survey %>%
  semi_join(shared_transects) %>%
  group_by(site, date, plant.sp) %>% # group by plant.sp*transect
  summarize(fl.abund = sum(flower.cover)) %>% # get plant.sp*transect total flower cover; this step really shouldn't be required, but in a few cases, there is more than one abundance entry for the same species on a given transect; not sure how that happened, but it's in the original data that Fabrice sent me; alternatively, I could drop the dups or average them. We'll see what Kalli and Fabrice have to say.
  group_by(site, date) %>%
  mutate(prop.fl.abund = fl.abund/sum(fl.abund)) %>% # get the proportional flower cover for each species.
  ungroup() %>%
  full_join(dplyr::select(net, site, date, plant.sp), by = c("site", "date", "plant.sp")) %>% # add the handful of species occurring in the visitation network but missing from the survey data
  complete(site, date, plant.sp, fill = list(fl.abund = 0, prop.fl.abund = 0)) %>%
  unique()
```

### Get most important plants for each BB species
```{r}
# All possible plant-bb combos
fl_bb_pair_filler <- full_join(bb_abund_sp, fl_abund_sp) %>%
  dplyr::select(site, date, bb.sp, plant.sp) %>%
  complete(site, date, bb.sp, plant.sp)

fl_bb_pair_filter_sub <- fl_bb_pair_filler %>%
  right_join(filter())

filter(fl_bb_all, is.na(site))
filter(fl_bb_all, is.na(date))
filter(fl_bb_all, is.na(bb.sp))
filter(fl_bb_all, is.na(plant.sp))
filter(fl_bb_all, plant.sp == "Rubus idaeus")

fl_abund_sp_unique <- unique(fl_abund_sp)

# This creates a data set where each bumble bee species present at a given site*date is paired with each floral species present at a given site date, even if there was never a visitation between them.
observed_visits <- net %>%
  semi_join(shared_transects) %>% # include only site*date combos where floral surveying was done
  group_by(site, date, bb.sp, plant.sp) %>% # group by bb*plant*transect
  summarize(visits = n()) %>% # total visits per bb-fl pair
  group_by(site, date, bb.sp) %>%
  mutate(prop.visits = visits/sum(visits))

# All possible plants * site * date
possible_plants <- bind_rows(dplyr::select(net, c(site, date, plant.sp)),
                             dplyr::select(survey, c(site, date, plant.sp))) %>%
  unique() %>%
  complete(site, date, plant.sp)

# All possible bumbles * site * date
possible_bb <- net %>%
  dplyr::select(site, date, bb.sp) %>%
  unique() %>%
  complete(site, date, bb.sp)

# All possible bumble-plant pairings *site * date, filtered to include only those when the given bb.sp was present and the given plant.sp was *either* present *or* visited (see note below)
visitation_by_abundance <- full_join(possible_plants, possible_bb) %>%
  left_join(bb_abund_sp) %>%
  left_join(fl_abund_sp) %>%
  #filter(fl.abund > 0) %>%
  filter(bb.abund > 0) %>%
  left_join(observed_visits) %>%
  filter(fl.abund > 0 | visits > 0) %>% # In about 10% of the bb.sp*plant.sp*site*year combos, the visited plant was not recorded in the floral survey data. There are a few different ways we could handle this, but I think an argument could be make that treating those plants as if they had zero abundance (even though they were evidently present gets close to the truth, since they must have been rather rare to have been missed in the surveying.)
  replace_na(list(visits = 0, prop.visits = 0))

write_csv(visitation_by_abundance, "../analyses/output/visitation_by_abundance.csv")

ggplot(visitation_by_abundance, aes(log(prop.fl.abund + 0.01), log(prop.visits + 0.01))) +
  geom_point() +
  geom_smooth(method = "lm")

### Get mean abundance/visitation for each plant*bb combo
visitation_by_abundance_mean <- visitation_by_abundance %>%
  mutate(year = factor(year(date))) %>%
  #group_by(year, plant.sp, bb.sp) %>%
  group_by(plant.sp, bb.sp) %>%
  summarize(prop.fl.abund.m = mean(prop.fl.abund),
            prop.visits.wm = weighted.mean(prop.visits, bb.abund)) %>%
  mutate(selection.ratio = prop.visits.wm/prop.fl.abund.m)

write_csv(visitation_by_abundance_mean, "../analyses/output/visitation_by_abundance_mean.csv")

ggplot(visitation_by_abundance_mean, aes(selection.ratio)) +
  geom_histogram()

filter(visitation_by_abundance_mean, plant.sp == "Trifolium pratense")

ggplot(visitation_by_abundance_mean, 
       aes(log(prop.fl.abund.m + 0.01), log(prop.visits.wm + 0.01))) +
  geom_point() +
  geom_smooth(method = "lm")

ggplot(filter(visitation_by_abundance_mean, prop.visits.wm > 0), 
       aes(log(prop.fl.abund.m + 0.01), log(prop.visits.wm + 0.01))) +
  geom_point() +
  geom_smooth(method = "lm")
```

### Okay, here's what we're going to do. Preference has a strong bimodal pattern, with lots of plant species totally ignored while the ones that are foraged upon are visited in proportion to their abundance. We can't model that latter pattern well if we retain all the totally ignored plant species. So, let's model the visited plants separately from the unvisited ones. 
```{r}
### Call linear model on the subset of the floral species that were visited at least once
pref.lm <- lm(log(prop.visits.wm) ~ log(prop.fl.abund.m + 0.01), # log on log regression is common for proportional data
              data = filter(visitation_by_abundance_mean, prop.visits.wm > 0))

@@

@@

Join residuals back into data and summarize by site-date-bb.sp (this deserves its own code chunk)
```{r}
#################################
### Build preference data set ###
#################################
# Start with the mean prop.fl.abund and weighted mean prop.visits
preference <- visitation_by_abundance_mean %>%  
  ungroup() %>%

@@

# Now we need to add back all the bb*fl*year combos that had 0 visits
# Then we have NAs for preference and floral abundance columns
  full_join(unique(dplyr::select(visitation_by_abundance_mean, bb.sp, plant.sp))) %>%

@@

# Drop the columns used to calculate preference so that they
# don't confuse us when we add site-date level abundance data
  dplyr::select(bb.sp, plant.sp, preference)

@@

preference_weighted_abund <- fl_abund_sp %>%
  right_join(preference, by = c("plant.sp")) %>% 
  mutate(fl.abund.pref_wt = fl.abund*preference) %>%
  mutate(year = factor(year(date)),
         yday = yday(date)) %>%
  group_by(site, year, date, yday, bb.sp) %>%
  summarize(net.fl.abund.pref_wt = sum(fl.abund.pref_wt, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(site_data, by = "site")

ggplot(preference_weighted_abund, aes(net.fl.abund.pref_wt)) +
  geom_histogram()

ggplot(preference_weighted_abund, aes(log(net.fl.abund.pref_wt + 0.01))) +
  geom_histogram()
```

Preference-weighted floral abundance by year
```{r}
pref_sum <- preference_weighted_abund %>%
  group_by(site, year, bb.sp) %>%
  summarize(mean.net.fl.abund.pref_wt = mean(net.fl.abund.pref_wt))

ggplot(pref_sum, aes(year, mean.net.fl.abund.pref_wt, color = bb.sp)) +
  geom_point()

ggplot(pref_sum, aes(year, log(mean.net.fl.abund.pref_wt + 0.01))) +
  geom_boxplot() +
  facet_wrap(~bb.sp, scales = "free")
```

GAM modeling of preference-weighted abundance per bb species
## Against any potential objections to my use of an added constant to avoid the log(0) problem, I would argue that when what is being measured is a resource, the presence of a resource in vanishingly small amounts is biologically equivalent to its absence. I suppose the strength of this argument hinges on how much we trust that our samples represent our sites. But if they don't, what are we even doing here? note that this argument does not hold when one is concerned with diversity and species distributions, since then a species rarity is a very different thing that its absence.
```{r}
### Set constants
k <- min(
  filter(preference_weighted_abund, net.fl.abund.pref_wt > 0)$net.fl.abund.pref_wt
)

### Call models
weighted_fl_abundance.2010 <- bam(net.fl.abund.pref_wt ~ te(yday, elev.mean, 
                                                            by = factor(bb.sp),
                                                            bs= c("tp", "tp"), 
                                                            k=c(12, 12), m=1),
                        data = filter(preference_weighted_abund, year == 2010),
                        discrete = TRUE,
                        method = "fREML") %>% getViz()

@@

weighted_fl_abundance.2011 <- bam(log(net.fl.abund.pref_wt + k) ~ te(yday, elev.mean, 
                                                  by = factor(bb.sp),
                                                  bs= c("tp", "tp"), 
                                                  k=c(10, 10), m=1),
                        data = filter(preference_weighted_abund, year == 2011),
                        discrete = TRUE,
                        method = "fREML") %>% getViz()

@@

weighted_fl_abundance.2012 <- bam(log(net.fl.abund.pref_wt + k) ~ te(yday, elev.mean, 
                                                  by = factor(bb.sp),
                                                  bs= c("tp", "tp"), 
                                                  k=c(10, 10), m=1),
                        data = filter(preference_weighted_abund, year == 2012),
                        discrete = TRUE,
                        method = "fREML") %>% getViz()

### 2020-07-03 writing/manuscript_v3.Rmd
fixed some things, broke some things

---
title: Bumble bees and wildflowers along an elevational climate gradient exhibit niche
  asymmetry and non-linear abundance patterns
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter 
output:
  # pdf_document: default
  # word_document:
  #   reference_docx: docx_template.docx
bibliography: refs.bib
output: html_document
---

@@

**Methods**
*Study sites*
*Sampling*
*Growing degree day analysis*
*Diversity modeling*
- GDM
*Abundance modeling*
- Floral preference modeling
- Preference-weighted floral abundance
- GAM
*Network analysis*
- Bootstrapping
- Network metrics

>>>>

---
title: Bumble bees and wildflowers along an elevational climate gradient exhibit niche
  asymmetry and heterogeneous abundance patterns
author:
- Douglas B. Sponsler
- Fabrice Requier
- Katharina Kallnik
- Alice Claßen
- A. Fabienne Maihoff
- Ingolf Steffan-Dewenter 
output:
  # pdf_document: default
  # word_document:
  #   reference_docx: docx_template.docx
bibliography: refs.bib
output: html_document
---

@@

**Methods**
*Study sites*
*Sampling*
*Growing degree day analysis*
*Diversity modeling*
- GDM
*Abundance modeling*
- Floral preference modeling

Floral preference was estimated for all pairwise combinations of bumble bee and plant species.

- Preference-weighted floral abundance
- GAM
*Network analysis*
- Bootstrapping
- Network metrics
